---
title: "Tutorial for goric(): How to evaluate theory-based hypotheses using the GORIC
  and GORICA"
author: "Rebecca M. Kuiper"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  #pdf_document:
  #  toc: yes
  html_document:
    toc: yes
    toc_float: yes
    number_sections: yes
fontsize: 14pt
editor_options:
  chunk_output_type: console
---

<style>
body {
  width: 100%;
  margin: 0 auto;
}
</style>


```{r setup, include = FALSE}
knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 80), tidy = TRUE)
knitr::opts_chunk$set(comment = NA, warning = FALSE)
options(width = 1200) # width console output
```

# Introduction

## Analyses

Below, you will find examples for the use of the `goric()` function in the `restriktor`-package. 

Two types of analyses can be run: 

* `type = "goric"` (the default), which can be applied to multivariate normal linear models; 
* `type = "gorica"`, the approximation of the GORIC which can be applied to a broad range of models.


Furthermore, there is the argument `comparison =` for the following three types of comparisons: 

* `"none"`
* `"unconstrained"` (default)
* `"complement"`

If you choose `"none"`, only the hypotheses of interest are inspected. This can lead to choosing the best out of a set of weak hypotheses. Therefore, this is only recommended when the hypotheses of interest cover the full parameter space / cover the whole set of possible theories. 
If `"unconstrained"` is chosen, then the unconstrained / unrestricted / classical alternative hypothesis is included in the set. This safeguards from choosing a weak hypothesis, that is, a hypothesis not supported by the data, as the best one. If at least one of the theory-based / informative hypotheses is not weak, one can compare those to all other hypotheses in the set.
Currently, `"complement"` only works for one hypothesis and not for a whole set. Then, the complement of the hypothesis of interest is evaluated and acts like a competing hypothesis. Since there is per definition no overlap between the hypothesis and its complement, this is more powerful then including the unconstrained hypothesis.
Please note that in the example below, the default `comparison = "unconstrained"` is used. At the end of the file, there are three examples for the three types of comparisons.


The `goric()` function of `restriktor` takes different forms of input (for model parameter estimates and constraints). 

(@) Fitted unconstrained (`lm` or `glm`) object* + character constraints.
(@) Fitted unconstrained (`lm` or `glm`) object + list with constraints matrix.
    + In case of within effects, the GORIC cannot be calculated.
(@) Fitted `restriktor` object(s).
(@) Numeric vector + character constraints.
(@) Numeric vector + list with constraints matrix.

Please note that:

* The last two (i.e., using the numeric factor) can only be used when `type = "gorica"` (the first three can be used for both goric and gorica).
* Using the numeric factor can be especially helpful for objects other than (g)lm objects.
* The numeric vector contains parameter estimates, and these can be obtained from any type of model.
* The estimates are assumed to be normally distributed. So, for some models when sample size is low, this assumption may not hold. In that case, it is often not clear how well the GORICA performs. See Altinisik et al 2021, for some simulations regarding logistic regression and SEM models.


## Hypothesis specification

A few remarks on how to specify hypotheses when working with the `goric()` function. When one uses `character constraints`:

 * Within the `restriktor()` and `goric()` functions, it is possible to use the following operators: `>`, `<`, `=`, `<=`, `>=`, `==` (where the last three denote the same constraint as the first three). 

 * The `goric()` and the `restriktor()` functions can deal with:
      + pairwise restrictions separated by a semicolon `;` (e.g., *"beta1 > beta2; beta2 = beta3"*).
      + combined restrictions consisting of more than one operator (e.g., *"beta1 > beta2 = beta3"*).
  Note that one should use the labels of the parameter estimates.
      
 * One can also define hypothesis in terms of linear functions of parameters (For more details, see 'Extra possibility specification hypotheses' near the end of this document).

When one wish to use a list with constraints matrix (possibly together with *rhs* and *neq*), it should be done as follows:
 `list(constraints = xxx, rhs = xxx, neq = x)`, with `constraints`  being a matrix, `rhs` a vector with the right hand sie of the restrictions, where the length of the vector equals the number of rows of constraints matrix (when not all of them are zero), and `neq` a scalar denoting the number of equalities in the matrix (when not zero). By default, `rhs` and `neq` are set to 0.
These arguments have to be specified such that: 
      + `constraints * parameters >= rhs`,
      + where the first `neq`-restrictions are equalities.

In the following examples, the above methods are illustrated.


# Preparation

Before starting the analyses, it is necessary to install and load the packages. Note that one needs to install the package only once, but has to load the package each time the R script is run.

```{r, message=FALSE, warning=FALSE}
# If you want to use restriktor from github:
#if (!require("devtools")) install.packages("devtools")
#library(devtools) 
#install_github("LeonardV/restriktor")
#library(restriktor) # for goric function
#
# If from CRAN:
if (!require("restriktor")) install.packages("restriktor") 
library(restriktor) # for goric function
```

Next, data is generated as example data to illustrate the functionality of the goric() function. 
If you want to obtain the same results as obtained here, it is necessary to set the seed at a fixed number.

```{r}
set.seed(123)

n <- 10        # sample size
x1 <- rnorm(n) # predictor 1, where n random draws are taken from a standard 
               #                               normal distributed population
x2 <- rnorm(n) # predictor 2
x3 <- rnorm(n) # predictor 3
e <- rnorm(n)  # error term
y <- 1 + x1 + x2 + x3 + e # outcome, obtained by filling in the regression 
                          # equation with an intercept of 1 and 
                          # regression coefficients of 1.

```

Then, a regression model can be fit using the `lm()` function:

```{r}
fit.lm <- lm(y ~ x1 + x2 + x3)
```

Below, the GORIC and GORICA are applied to this model. 

To apply the GORIC(A), hypotheses need to be specified. Normally, these would be based on theory. However, since (hypothetical) data is generated, quasi-randomly two hypotheses are specified. 

Notes on specifying hypotheses:

* Hypotheses in R are in terms of the labeling used for the estimates (instead of the names for the population parameters, say, beta0, ..., beta3). Thus, here: .Intercept., x1, x2, and/or x3. 

* In case estimates of continuous predictor variables are compared (e.g., `"x1 < x2"`), one should standardize the data (this is shown in more detail below) or use the standarized parameter estimates.


# Model Selection using GORIC(A)

## GORIC & GORICA

The upcoming analyses can be performed by means of both the GORIC and the GORICA. By default, the `goric()` function calculates the **GORIC** values (which then uses: type = "goric"). To calculate the **GORICA** values, the argument `type` has to be set to `gorica` (i.e., type = "gorica") within the `goric()` function. 


Note on GORIC and GORICA weights:

* The GORICA weights (asymptotically) equal the GORIC weights. The differences are minor and often not notable with 2 decimals. Because of these minor differences, the relative weights (i.e., ratio of weights) can differ. These differences in relative weights can even be large, when dividing a very large number by a very small number with minor differences in these values.


### Fitted unconstrained (g)lm object + character constraints

Next, the hypotheses need to be specified.
Note that the `;` separates the restrictions within one hypothesis.

```{r}
H1 <- "x1 > 0"
H2 <- "x1 > 0; x2 > 0"
```

Now, the GORIC values and weights can be calculated.

```{r}
set.seed(123) # Set seed value
out <- goric(fit.lm, hypotheses = list(H1, H2))
summary(out)
```

To obtain the GORICA values and weights instead:
```{r}
set.seed(123) # Set seed value
out_gorica <- goric(fit.lm, hypotheses = list(H1, H2), type = "gorica")
summary(out_gorica)
```


### Fitted unconstrained (g)lm object + list with constraints matrix

In the example above, `rhs` and `neq` are assumed to be 0, which is also the assumed default. It means that the left hand side of (in)equalities are set to $0$'s, and that the number of equality restrictions is equal to $0$. 

```{r}
H1 <- list(constraints = c(0,1,0,0)) 
# "x1 > 0"
H2 <- list(constraints = rbind(c(0,1,0,0), c(0,0,1,0)))  
# "x1 > 0; x2 > 0" 
```

The GORIC values and weights are obtained via:

```{r}
set.seed(123) # Set seed value
out <- goric(fit.lm, hypotheses = list(H1, H2))
summary(out)
```

and the GORICA values and weights via:

```{r}
set.seed(123) # Set seed value
out_gorica <- goric(fit.lm, hypotheses = list(H1, H2), type = "gorica")
summary(out_gorica)
```

### Fitted unconstrained (g)lm object + list with constraints matrix (rhs and neq non-zero)

In this example, `rhs` and `neq` are not assumed to be 0. Please note that equality restrictions should be stated first in the constraints' list.

```{r}
H1 <- list(constraints = c(0,1,0,0), rhs = 2) 
# "x1 > 2"
H2 <- list(constraints = rbind(c(0,1,0,0), c(0,0,1,0)), rhs = c(1,0), neq = 1) 
# "x1 = 1; x2 > 0"
```

The GORIC values and weights are obtained via:

```{r}
set.seed(123) # Set seed value
out <- goric(fit.lm, hypotheses = list(H1, H2))
summary(out)
```

and the GORICA values and weights via:

```{r}
set.seed(123) # Set seed value
out_gorica <- goric(fit.lm, hypotheses = list(H1, H2), type = "gorica")
summary(out_gorica)
```

<!-- TO DO dit kan niet meer?? iig foutmelding... -->
<!-- ### Fitted restriktor object(s) -->

<!-- After specifying the hypotheses: -->

<!-- ```{r} -->
<!-- H1 <- "x1 > 0" -->
<!-- H2 <- "x1 > 0; x2 > 0" -->
<!-- ``` -->

<!-- the `restriktor` objects can be fitted as follows: -->

<!-- ```{r} -->
<!-- fit1.restr <- restriktor(fit.lm, constraints = H1) -->
<!-- fit2.restr <- restriktor(fit.lm, constraints = H2) -->
<!-- ``` -->

<!-- Now, use the GORIC for model selection: -->

<!-- ```{r} -->
<!-- set.seed(123) # Set seed value -->
<!-- out <- goric(fit1.restr, fit2.restr) -->
<!-- summary(out) -->
<!-- ``` -->

<!-- or the GORICA: -->

<!-- ```{r} -->
<!-- set.seed(123) # Set seed value -->
<!-- out_gorica <- goric(fit1.restr, fit2.restr, type = "gorica") -->
<!-- summary(out) -->
<!-- ``` -->


## GORICA

The upcoming analyses can only be performed by means of the *GORICA*. This is because, the GORICA only needs the parameter estimates or only those of interest (that is, those addressed in the hypotheses) and their covariance matrix. Because of this, it is applicable for a wider range of models (e.g., logistic regression and SEM models).


### Numeric vector + character constraints

One needs to extract the estimates (of interest), as well as their variance-covariance matrix. Both can be extracted from the `fit.lm` object (also from non-(g)lm objects).

```{r}
est <- coef(fit.lm)
VCOV <- vcov(fit.lm)
```

Next, the hypotheses need to be specified using the labels of 'est' (which can also be re-labelled as is done in the next subsection).

```{r}
H1 <- "x1 > 0"
H2 <- "x1 > 0; x2 > 0"
```

Then, use the `goric()` function to select the best model.

```{r}
set.seed(123) # Set seed value
out <- goric(est, hypotheses = list(H1, H2), VCOV = VCOV, type = "gorica")
summary(out)
```

However, one could also use the structural parameters only, that is, the parameters used in all hypotheses in the set. In this case, the following code can be used:

```{r}
H1 <- "x1 > 0"
H2 <- "x1 > 0; x2 > 0"

est <- coef(fit.lm)[2:3]
VCOV <- vcov(fit.lm)[2:3, 2:3]

set.seed(123) # Set seed value
out <- goric(est, hypotheses = list(H1, H2), VCOV = VCOV, type = "gorica")
summary(out)
```

Now, the exact same weights are obtained. However, the penalties are two points lower than before, which is due to the fact that two unconstrained/free parameters are left out. The log-likelihoods also change. But, when comparing the support for the hypotheses, these differences cancel out.


### Relabeled numeric vector + character constraints

By default, the names/labels of the parameter estimates are based on variable names in the model (and in case of a factor, it is also based on the different levels it can take on).
Using other names/labels often makes more sense. Then, one can specify the hypotheses in terms of those labels. For example, one can do the following:

```{r}
est <- coef(fit.lm)
names(est) <- c("beta0", "beta1", "beta2", "beta3")
VCOV <- vcov(fit.lm)

H1 <- "beta1 > 0"
H2 <- "beta1 > 0; beta2 > 0"

set.seed(123) # Set seed value
out <- goric(est, hypotheses = list(H1, H2), VCOV = VCOV, type = "gorica")
summary(out)
```

### 'Manual' numeric vector + character constraints

Sometimes, one cannot extract the coefficients and variance-covariance matrix directly from an R-object; for example, due to the fact that previous analyses were performed by means of other software, or since it stems from another article. In these situations, it is also possible to manually insert the regression coefficients and the variance-covariance matrix.

```{r}
est <- c(1.2479164, 1.4067410, 0.8260098, 0.8113972)

names(est) <- c("beta0", "beta1", "beta2", "beta3")

VCOV <- matrix(c(0.023513875, 0.002584345, 0.000504821, 0.011663312,
                 0.002584345, 0.035044625, -0.016490653, 0.004143737,
                 0.000504821, -0.016490653, 0.036437680, 0.016195388,
                 0.011663312, 0.004143737, 0.016195388, 0.036158130),
               byrow = TRUE, ncol = 4)

H1 <- "beta1 > 0"
H2 <- "beta1 > 0; beta2 > 0"

set.seed(123) # Set seed value
out <- goric(est, hypotheses = list(H1, H2), VCOV = VCOV, type = "gorica")
summary(out)
```

#Numeric vector + list with constraints matrix (rhs, neq)

When one only wants to use a list with a constraints matrix, the following code can be used:

```{r}
est <- coef(fit.lm)
VCOV <- vcov(fit.lm)

H1 <- list(constraints = c(0,1,0,0)) 
# "x1 > 0"
H2 <- list(constraints = rbind(c(0,1,0,0), c(0,0,1,0))) 
# "x1 > 0; x2 > 0"

set.seed(123) # Set seed value
out <- goric(est, hypotheses = list(H1, H2), VCOV = VCOV, type = "gorica")
summary(out)
```


## Standardized data

If estimates of continuous predictor variables are compared (e.g., *"x1 < x2"*), then one should standardize the data.
To demonstrate this, new data will be generated. In the data, the contribution of the variables to the dependent variable will differ. Below, the vector of ratios of these contributions is set to $=c(1,1.5,2)$. This, then reflects (the ratios of) the population regression coefficient values. It means that the contribution to the outcome variable `y`, increases from `x1` to `x3`, where that of `x2` is $1.5$ and that of `x3` is $2$ times as large as that of `x1`.

```{r}
ratio <- c(1, 1.5, 2) # ratio between population effects of the predictors

n <- 30 # sample size

x1 <- rnorm(n) # predictor1: 
               # n draws from a standard normal distributed population
x2 <- rnorm(n) # predictor2
x3 <- rnorm(n) # predictor3
e <- rnorm(n)

data <- cbind(x1, x2, x3) 
data_s <- as.data.frame(scale(data))

y <- ratio[1] * data_s$x1 + ratio[2] * data_s$x2 + ratio[3] * data_s$x3 + e
```

*Note.* Since there is only one outcome, the outcome does not need to be standardized.

Now, fit the regression model, and specify a hypothesis by means of character constraints.

```{r}
fit.lm_s <- lm(y ~ -1 + x1 + x2 + x3, data = data_s)

H1 <- "x1 < x2"
```

Then, calculate the GORIC values and weights:

```{r}
set.seed(123) # Set seed value
out_s <- goric(fit.lm, hypotheses = list(H1), comparison = "complement")
summary(out_s, brief = FALSE)
```


## Defining parameters

In `restriktor`, it is also possible to specify hypotheses in terms of linear functions of the parameters. For instance, in case of adjusted means, you can specify the restrictions in two ways:

1. You can write them out yourself (using the parameter/variable names from an R-object, like `lm()`).
2. You can first specify the adjusted means (below called `"m1"` to `"m4"`) and specify the hypotheses in terms of those:

<!-- ```{r} -->
<!-- H1 <- ' m1 := .Intercept. -->
<!--         m2 := .Intercept. + facialBurns -->
<!--         m3 := .Intercept. + gender -->
<!--         m4 := .Intercept. + facialBurns + gender + gender:facialBurns -->
<!--         m1 < m4 -->
<!--         m2 < m4 -->
<!--         m3 < m4 ' -->
<!-- ``` -->
<!-- Is nu: -->
```{r}
H1 <- ' m1 := .Intercept.
        m2 := .Intercept. + facialBurns
        m3 := .Intercept. + gender
        m4 := .Intercept. + facialBurns + gender + gender.facialBurns
        m1 < m4
        m2 < m4
        m3 < m4 '
```
For more details, please see Vanbrabant, Van Loey, and Kuiper 2020). However, please note that this code cannot be run now, since the variables and, more importantly, the parameters are labeled differently. One can run:

```{r}
H1 <- ' m1 := .Intercept.
        m2 := .Intercept. + x1
        m3 := .Intercept. + x2
        m4 := .Intercept. + x3
        m1 < m4
        m2 < m4
        m3 < m4 '

set.seed(123) # Set seed value
out <- goric(fit.lm, hypotheses = list(H1), 
             comparison = 'complement') # for info, see below
summary(out)
```


# Types of comparisons

There are three types of comparisons: `"none"`, `"unconstrained"`, and `"complement"`. Subsequently, it is shown, for the last type, what output can be asked for, which of course holds for all examples.


## Comparison = "none""

If `comparison = "none"`, then only the hypotheses of interest are inspected. This can lead to choosing the best hypothesis out of a set of weak hypotheses.
Hence, one should only use this if the hypotheses cover the whole parameter space, that is, cover all possible theories/hypotheses.

```{r}
H1 <- "x1 > 0"
H2 <- "x1 > 0; x2 > 0"

set.seed(123) # Set seed value
out <- goric(fit.lm, hypotheses = list(H1, H2), comparison = "none")
summary(out)
```

## Comparison = "unconstrained"

If `comparison = "unconstrained"`, then the unconstrained / unrestricted / classical alternative hypothesis is included in the set. This safeguards from choosing a weak hypothesis as the best one.

```{r}
H1 <- "x1 > 0"
H2 <- "x1 > 0; x2 > 0"

set.seed(123) # Set seed value
out <- goric(fit.lm, hypotheses = list(H1, H2), comparison = "unconstrained")
# or: 
# out <- goric(fit.lm, H1, H2, type = "gorica", comparison = "unconstrained")
summary(out)
```
Since at least one of, even both, H1 and H2 are not weak, one can compare the support for H1 versus H2 (where H2 is 1.921 times more supported).


## Comparison = "complement"

Currently, `comparison = "complement"` only works for one hypothesis and not for a whole set of hypotheses. 
Then, the complement of the hypothesis of interest is evaluated and acts like a competing hypothesis. Since there is per definition no overlap between the hypothesis and its complement, this is more powerful then including the unconstrained.


### H1 vs its complement

```{r}
H1 <- "x1 > 0"

set.seed(123) # Set seed value
out <- goric(fit.lm, hypotheses = list(H1), comparison = "complement")
summary(out, brief = FALSE)
summary(out)
```

### H2 vs its complement

```{r}
H2 <- "x1 > 0; x2 > 0"

set.seed(123) # Set seed value
out <- goric(fit.lm, hypotheses = list(H2), comparison = "complement")
summary(out)
```

### Note on comparisons

Please note that the obtained weights for the two hypotheses cannot be compared. If this is the goal of the analysis, one should evaluate them simultaneously in one set, as done earlier.



# Different types of output

The command `out` gives the default results (i.e., per model / hypothesis the log-likelihood, penalty, GORIC(A), and GORIC(A)-weights).

```{r}
out
```

The command `summary(out)` gives, besides the results, also the relative GORIC(A)-weights matrix.

```{r}
summary(out)
```

The command `summary(out, brief = FALSE)` gives also the order / inequality restricted coefficients (order-restricted mle's) and constraint matrices.

```{r}
summary(out, brief = FALSE)
```


Furthermore, one can ask for different elements of the output by means of the following commands:

* `out$result` - gives results (i.e., per model / hypothesis the log-likelihood, penalty, GORIC(A), and GORIC(A)-weights).
* `out$ratio.gw` - gives the relative GORIC(A)-weights matrix, which gives the relative support of an hypothesis versus another.
* `out$ormle` - gives the order / inequality restricted coefficients (order-restricted mle's).
* `coef(out)` - also gives the order / inequality restricted coefficients (order-restricted mle's).
* `out$type` - states what type of analysis is used (GORIC or GORICA).
* `out$comparison` - states what type of comparison is used (none, unconstrained, complement).
* `out$constraints` - gives the constraints matrices.
* `out$rhs` - gives the right hand side (rhs) of the constraints.
* `out$neq` - gives the number of equalities (neq) in the constraints, where the first neq are then equalities.


# If many restrictions: Another method to calculate the penalty

Working with a lot of restrictions and a lot of parameters calculations increases the computation time, which has to do with the calculation of the level probability needed for the penalty term. 
There are two methods that can be used in calculating the penalty. The default method is often much faster (if the number of parameters is not too high) and needs less input specification. It can, however, not deal with hypotheses that are not of full row-rank (which easily happens if you specify many restrictions). In that case, `restriktor` uses automatically the other (bootstrap) method. 

To use this bootstrap method, use:
```{r, message = FALSE, warning = FALSE}
#if (!require("parallel")) install.packages("parallel") 
#library(parallel)
#nrCPUcores <- detectCores(all.tests = FALSE, logical = TRUE)
```
```{r}
set.seed(123) # Set seed value
goric(est, VCOV = VCOV, hypotheses = list(H1), comparison = "complement",
      mix_weights = "boot")
```


<!-- In case of bootstrapping, the following additional arguments are used: -->

<!-- * `mix_weights` - method for calculating the level probabilities (options: "none", "pmvnorm", "boot"). -->
<!-- * `convergence_crit ` - number of bootstrap samples. -->
<!-- * `chunk_size ` - use of parallel processing. Windows: parallel = 'snow'; unix: parallel = 'multicore'. -->
<!-- * `ncpus` - number of processors in your PC. -->

<!-- This can also be helpful in case the restriction matrix is not full row-rank, as discussed next.  -->


# Note on not full row-rank

If the restriction matrix is not of full row-rank, this means one of the following:

1. Redundant restriction, either:
    a) leave the redundant one out (goric() function does this for you), or:
    b) use another (more time-consuming) way of obtaining the level probabilities for the penalty term: Bootstrapping, as discussed above (goric function does this by default). 
2. Range restrictions (e.g., -2 < group1 < 2), which can be evaluated, but there is a sensitivity (like with a prior in a Bayes factor), which currently cannot be checked for.
3. Conflicting restrictions (e.g., 2 < group1 < -2), which can evidently never hold and is thus impossible to evaluate. In this case, one should delete the incorrect restriction.



[//]: #The following line is needed to prevent R Markdown from including a lot of white space below the last content.
<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 0;"></div>