---
title: 'Evaluate a set of hypotheses with GORIC or GORICA: ANOVA Example'
author: "Rebecca M. Kuiper"
date: "19-7-2021"
output:
  html_document:
    toc: yes
    toc_float: yes
  pdf_document:
    toc: yes
  word_document:
    toc: yes
fontsize: 14pt
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(tidy.opts = list(width.cutoff = 80), tidy = TRUE)
knitr::opts_chunk$set(comment = NA,warning=FALSE)
```

# Example 1: Lucas Data

## Data preparation

First, we need to load the required libraries (after they have been installed). These libraries contain functions, such as `goric`, that will be used in the R code below.
Each time you reopen R, you will have to load the required libraries.

```{r, message = FALSE, warning = FALSE}
## First, we need to install the packages, if you have not done this already:
# If you want to use restriktor from github:
#if (!require("devtools")) install.packages("devtools")
#library(devtools) 
#install_github("LeonardV/restriktor")
#library(restriktor) # for goric function

## First, we need to install the packages, if you have not done this already:
if (!require("psych")) install.packages("psych")
if (!require("restriktor")) install.packages("restriktor")

## Then, load the packages:
library(psych) # for the function describeBy
library(restriktor) # for the goric function
```


Second, it is necessary to load the data.

Notably, it is only possible to load the data if you are using the correct working directory (with both your R script and data file). The command `getwd()` shows you your current working directory. You can change the working directory to the one you prefer using the function `setwd()` by specifying the correct location between parentheses. Alternatively, in Rstudio, you can use the "Session" tab (on top) or you can use the "Files"-pane (on top of probably the right lower box of your Rstudio-screen, this pane is located next to the panes for "Plots", "Packages", "Help" and "Viewer").

If you open the data file `Data_Lucas.txt` in a text editor, you can see that the variable labels have been inserted (using quotes; i.e., "...") in the first line of the file, which is called a header. Therefore, you have to specify 'header = TRUE' when loading the data:

```{r}
# Load the data
Lucas <- read.table("Data_Lucas.txt", header = TRUE)
```

Since we loaded a .txt file, R does not know the measurement levels of the variables and assumes all of them to be continuous, meaning that they are of interval or ratio type. Hence, we need to tell R that the variable `group` is a factor by using the `factor()` function over the `group` variable (i.e., a grouping / categorical / nominal variable): 

```{r}
# Make the variable group a factor
Lucas$group <- factor(Lucas$group) 
```

To inspect the first 6 rows of the dataset, we use `head()` function:

```{r}
head(Lucas) # Look at first (6) rows of the data
```

To see a more detailed overview of the data via descriptive statistics split by `group` variable, we could use the `describeBy()` function with `Lucas$group` set to be a grouping variable, as follows:

```{r}
descrstat <- describeBy(Lucas$Influence, Lucas$group, mat = TRUE)
descrstat
```


## Preparation for GORIC(A)

#### ANOVA model: R-object

First, we need an R-object with unconstrained estimates, that is, in this example, the five group means and one residual variance.We specify the linear regression model using `lm()` function, as follows:

```{r}
lm_fit_Lucas <- lm(Influence ~ group - 1, data = Lucas)
```

Note that:

  1. `y ~ group - 1` instructs the function `lm` (linear model) to regress y on group variable.
  
  2. The `- 1` instructs the function `lm` to drop the intercept and, therefore, estimate the means of each group, resulting   here in five group means. 
On the other hand, if we wouldn't drop the intercept `y ~ group' would estimate an intercept, representing the mean of the reference group, and the mean differences between the other (here, four) groups and the reference group.
  
  3. The results are collected in, what is called, an R-object, named `lm_fit_Lucas`.



Let's check the names used in this model.

```{r}
names(coef(lm_fit_Lucas))
```
  
  
#### ANOVA model: coef() and vcov()
Note that goric() can use lm or glm objects (even most lavaan objects) as input.
Since the GORICA can be applied to a broad range of models, the GORICA has another input option as well:
One can enter the (structural) parameter estimates and their covariance matrix. 
For most R-objects, these can be obtained using coef() and vcov().

E.g., 
```{r}
est <- coef(lm_fit_Lucas)
VCOV <- vcov(lm_fit_Lucas)
```

This will be illustrated in an GORICA example below.


 
#### Set of hypotheses
To evaluate the hypotheses of interest, it is necessary to specify the restrictions in these hypotheses correctly. Here are some points that user has to take into account during the procedure:  

 * It is possible to use the following operators: `>`, `<`, `=`, `== `,`<=`or`>=` within the `restriktor()` and `goric()` functions 
      + `==` operator is interpreted in the same fashion as the `=`, meaning an equality  
      + `<=` and `>=` operators are interpreted as respectively: `<` and `>` by the code

 * The `goric()` and the `restriktor` functions can deal with:
      + pairwise restrictions (eg. "x1>x2;x2==x3" also equivalent to "x1>x2;x2=x3")
      + combined with more than one operators restrictions(eg. "x1>x2==x3" also equivalent to "x1>x2=x3")
      
It is important to remember that all restrictions within one hypothesis has to be separated with a semicolon `;`.

In the examples below, 'full output' for the `goric()` function will be shown. At the of the tutorial, in the last example for the three types of comparisons, we will show how one can ask the function for specific output parts.

Let's specify hypothesis first:
```{r}
H0 <- 'group1 == group2 == group3 == group4 == group5' #'group1 = group2 = group3 = group4 = group5'
                                                       # could also be used 
H1 <- 'group5 == group3 > group1> group2; group3 > group4 > group2' 
# Note: H1 is not full row-rank(see below and the goric() tutorial for more details).

H2 <- 'group3 > group1 > group4 == group5 > group2'
# Note: 
#H2 could also be specified as pairs 'group3 > group1; group1 > group4; group4 == group5; group5 > group2'
```

To prevent from selecting a weak hypothesis, one should include a failsafe/safeguard hypothesis.
This can be:

  *   the unconstrained hypothesis (which includes all possible hypotheses, thus including the one(s) of interest);

  *   the complement (which includes all other possible hypotheses, thus excluding the one(s) of interest).
  
Currently, the complement can only be used for one hypothesis of interest. Therefore, the examples w.r.t the use of the complement only evaluate H1 (and not the whole set).


#### Seed values
In the calculation of the GORIC, an iterative process is used to calculate the penalty / complexity part. Therefore, one needs to set a seed value using the `set.seed()`. 
This will allow us to:

1. Reproduce outcomes for the penalty( we will obtain the same penalty value every time we run this code).

2. Access the sensitivity of the penalty value by checking for different seed values. If it is sensitive, then increase number of iterations used in calculation of the penalty (see below).


## GORIC examples

Keep in mind that the `goric()` function is calculating the *GORIC* value by default(type = "goric"). To calculate the *GORICA* values the argument `type` has to be set to `gorica`(see example below).

The GORIC can be used for normal linear models, like ANOVA and regression models.


### Example 1.1a: Using the unconstrained as failsafe
```{r}
set.seed(123) # Set seed value
output <- goric(lm_fit_Lucas, H0, H1, H2)
summary(output)
```

It can be seen that the order-restricted hypothesis $H_1$ has 16.5 times more support than $H_u$ (the unconstrained hypothesis). Hence, $H_1$ is not a weak hypotheses and can thus be compared to the other (weak and non-weak) competing hypotheses: $H_1$ is much more (4.52e+05 times more) supported than $H_0$ and 38.5 times more likely than $H_2$.


### Example 1.1b: Using the complement as failsafe 
```{r}
set.seed(123) # Set seed value
output_c <- goric(lm_fit_Lucas, H1, comparison = "complement")
summary(output_c)
```

The order-restricted hypothesis $H_1$ has 13.4 times more support than its complement.



## GORICA examples

Coming to calculating the *GORICA* values, we set `type = "gorica"`.

Note, that the *GORICA* can be used for a broad range of models. Besides normal linear models (e.g., ANOVA and regression models) it can be applied also to SEM models and much more.


### Example 1.2a: Using the unconstrained as failsafe
```{r}
set.seed(123) # Set seed value
output_gorica <- goric(lm_fit_Lucas, H0, H1, H2, type = "gorica")
summary(output_gorica)
```

In Example 1, the same analysis is done with the GORIC, you can see that the (relative) weights are (about) the same for the GORIC and GORICA.

From this output, it can be seen that the order-restricted hypothesis $H_1$ has 16.5 times more support than $H_u$ (the unconstrained hypothesis). Hence, $H_1$ is not a weak hypotheses and can thus be compared to the other (weak and non-weak) competing hypotheses: $H_1$ is much more (1.13e+06 times more) supported than $H_0$ and 37.2 times more likely than $H_2$.

#### Alternative input
```{r}
set.seed(123) # Set seed value
est <- coef(lm_fit_Lucas)
VCOV <- vcov(lm_fit_Lucas)
output_gorica_alt <- goric(est, VCOV = VCOV, H0, H1, H2, type = "gorica")
summary(output_gorica_alt)
```


### Example 1.2b: Using the complement as failsafe
```{r}
H1 <- 'group5 == group3 > group1 > group2; group3 > group4 > group2' # Note: H1 is not full row-rank, see the goric tutorial for more details.

set.seed(123) # Set seed value
output_gorica_c <- goric(lm_fit_Lucas, H1, comparison = "complement", type = "gorica")
summary(output_gorica_c)
```
The order-restricted hypothesis $H_1$ has 13.4 times more support than its complement.
Notably, the weights are also now the same as for the GORIC (Example 2).




# Example 2: Berzonsky et al.
Next, we present the R code to apply the GORIC(A) to the data of Berzonsky et al.. The instruction on loading and preparing the data are similar to the ones from Lucas example (Example 1). Hence, for more comments and details look up that example. 

## Data preparation
```{r}
# Read Data. 
BerzEtAl <- read.table("Data_BerzEtAl.txt", header = TRUE)
BerzEtAl$group <- factor(BerzEtAl$group) # this command tells R that group is a factor and not a continuous variable

# Inspect data
head(BerzEtAl)

# Compute descriptive statistics for each group
descrip <- describeBy(BerzEtAl$Influence, BerzEtAl$group, mat = TRUE)
descrip
```


## Preparation for GORIC(A)
```{r}
# Using the R package lm
lm_fit_BerzEtAl <-  lm(Influence ~ group-1, data=BerzEtAl)

# Check names used in model
names(coef(lm_fit_BerzEtAl))
# Specify restrictions using those names
```

  
 
#### Set of hypotheses
```{r}
H0 <- 'group1 == group2 == group3 == group4 == group5 == group6 == group7 == group8' 

H1 <- 'group1 > group2; group1 > group3; group1 > group4; 
      group5 > group6; group5 > group7; group5 > group8;  
      group1 > group5; group2 > group6; group3 > group7; group4 > group8;
      group1 - group5 > group2 - group6; group1 - group5 > group3 - group7; group1 - group5 > group4 - group8;'

H2 <- 'group1 > group2; group1 > group3; group1 > group4; 
      group5 > group6; group5 > group7; group5 > group8;  
      group1 > group5; group2 > group6; group3 > group7; group4 > group8;
      group1 - group5 > 2*(group2 - group6); group1 - group5 > 2*(group3 - group7); group1 - group5 > 2*(group4 - group8);'
```



## GORIC examples

### Example 2.1a: Using the unconstrained as failsafe
```{r}
set.seed(123) # Set seed value
output_B <- goric(lm_fit_BerzEtAl, H0, H1, H2) 
summary(output_B)
```
The order-restricted hypothesis $H_1$ has  17.4 times more support than $H_u$ (unconstrained).
Hence, $H_1$ is not a weak hypotheses and can thus be compared to the other (weak and non-weak) competing hypotheses: 
$H_1$ is much more (8.55e+56 and 4.73e+10 times more) supported than $H_0$ and $H_2$ (respectively).


### Example 2.1b: Using the complement as failsafe 
```{r}
set.seed(123) 
output_B_c <- goric(lm_fit_BerzEtAl, H1, comparison = "complement") 
summary(output_B_c)
```
The order-restricted hypothesis $H_1$ has 17.2 times more support than its complement.


## GORICA examples

### Example 2.2a: Using the unconstrained as failsafe
```{r}
set.seed(123) # Set seed value
output_B_gorica <- goric(lm_fit_BerzEtAl, H0, H1, H2, type = "gorica") 
summary(output_B_gorica)
```
The order-restricted hypothesis $H_1$ has  18.6 times more support than $H_u$ (unconstrained).
Hence, $H_1$ is not a weak hypotheses and can thus be compared to the other (weak and non-weak) competing hypotheses: 
$H_1$ is much more (3.68e+205 and 4.46e+12 times more) supported than $H_0$ and $H_2$ (respectively).



### Example 2.2b: Using the unconstrained as failsafe
```{r}
set.seed(123) 
output_B_gorica_c <- goric(lm_fit_BerzEtAl, H1, comparison = "complement", type = "gorica") 
summary(output_B_gorica_c)
```
The order-restricted hypothesis $H_1$ has 18.5 times more support than its complement.





# Notes on weights

Here, the example of Lucas is used again. We will use the following hypotheses of interest:
```{r}
H0 <- 'group1 == group2 == group3 == group4 == group5' 
H1 <- 'group5 == group3 > group1 > group2; group3 > group4 > group2' 
# Note: H1 is not full row-rank, see below and the goric tutorial for more details.
H2 <- 'group3 > group1 > group4 == group5 > group2'
```


### Note 1: GORIC weights versus GORICA weights

The GORICA weights equal the GORIC weights - the differences are minor and often not notable with 2 decimals.
Because of these minor differences, the relative weights (i.e., ratio of weights) can differ.
Those differences in relative weights can even be large (as is in the Berzonsky et al Example), when dividing a very large number by a very small number with minor differences in these values.



### Note 2: complement in case of two hypotheses of interest

One cannot compare the support of hypotheses when comparing them to their complements. Instead,
one should evaluate them simultaneously in one set to get desirable outcomes (like done above):

```{r}
# Calculate goric for H1 and its complement
set.seed(123) # Set seed value
output_c_H1 <- goric(lm_fit_Lucas, H1, comparison = "complement")
summary(output_c_H1)
# The order-restricted hypothesis H1 has  13.4 times more support than its complement.
#
# Calculate goric for H2 and its complement
set.seed(123) # Set seed value
output_c_H2 <- goric(lm_fit_Lucas, H2, comparison = "complement")
summary(output_c_H2)
#The order-restricted hypothesis H1 has  0.37 times more support than its complement.
#
# Calculate goric for H1 and H2 (and Hu):
set.seed(123) # Set seed value
output_H1H2 <- goric(lm_fit_Lucas, H1, H2) # Note: by default against the unconstrained
summary(output_H1H2)
# H1 has 38.5 times more support than H2.
# This is not equal to:
output_c_H1$relative.gw[1,2] / output_c_H2$relative.gw[1,2] # approx 13.4 / 0.37
# 36.34

# Notably, you could derive the support from H1 vs H2 from their support versus that of Hu:
set.seed(123) # Set seed value
output_u_H1 <- goric(lm_fit_Lucas, H1)
summary(output_u_H1)
#
set.seed(123) # Set seed value
output_u_H2 <- goric(lm_fit_Lucas, H2)
summary(output_u_H2)
#
output_u_H1$relative.gw[1,2] / output_u_H2$relative.gw[1,2] # approx (0.943/0.057) / (0.3/0.7) = 16.481 / 0.428
# 38.49661
# which (approximately) equals:
output_H1H2$relative.gw[1,2]
# 38.49665
```


### Note 3: weights when using complement not per se higher

In the Lucas example:

* The order-restricted hypothesis $H_1$ has  16.5 times more support than $H_u$ (unconstrained).
* The order-restricted hypothesis $H_1$ has  13.4 times more support than its complement $H_m$.

Now, the complement does not render higher weight than if $H_u$ was used.


If $H_m$ is not in agreement with the data, the complement does not always render a higher weight. 
This is actually a good thing: Against $H_u$, $H_m$ might obtain too much support then.


#### Some more explanation

This is because for this particular example the mean of group 3 and 5 are close (and they are compared in Hypothesis $H_1$). 
Notably, it will hold for all of the following three hypothesis:
```{r}
H1_gr <- 'group5 > group3 > group1 > group2; group3 > group4 > group2' # not full row-rank
H1_sm <- 'group5 < group3 > group1 > group2; group3 > group4 > group2' # not full row-rank
H1 <- 'group5 == group3 > group1 > group2; group3 > group4 > group2' # not full row-rank
```

Because the means of groups 3&5 are similar(and the other restrictions are in agreement with the data),
the log likelihood values (LL's) of $H_1$, $H_u$ and Hc (i.e., the complement of $H_1$) will be close - since the restricted/bounded solution (which is in agreement with $H_1$) is near the (unconstrained) maximum likelihood estimate (mle).
Since the penalty for the complement is (always) lower than for $H_u$, it will receive more support than $H_u$ - thus $H_1$ receives less.
Notably, when $H_1$ is very specific (as in Berzonsky et al Example below), the penalty of $H_u$ and Hcomplement are almost the same.

When the means of groups 3 and 5 differ more, then evaluating against the complement does render a higher weight (than if $H_u$ were used):
Let us increase the mean of group 5 with 0.5 points (and change the name of the data & the analysis):
```{r}
Lucas2 <- Lucas
Lucas2$Influence[Lucas$group == 5] <- Lucas$Influence[Lucas$group == 5] + 0.5
#describeBy(Lucas$Influence, Lucas$group, mat = TRUE)
#describeBy(Lucas2$Influence, Lucas$group, mat = TRUE)
lm_fit_Lucas2 <-  lm(Influence ~ group-1, data = Lucas2)
# Now, from the three hypotheses H1_gr, H1_sm, and H1, the first (H_gr) is correct, so that one is used to illustrate our point:
H1_gr <- 'group5 > group3 > group1 > group2; group3 > group4 > group2' # not full row-rank
set.seed(123) # Set seed value
output_u_gr <- goric(lm_fit_Lucas2, H1_gr)
summary(output_u_gr)
#The order-restricted hypothesis 'H1' has 13 times more support than Hu (unconstrained). Notably, this is the maximum support H1 can receive versus Hu (because LL's are the same).
set.seed(123) # Set seed value
output_c_gr <- goric(lm_fit_Lucas2, H1_gr, comparison = "complement")
summary(output_c_gr)
#The order-restricted hypothesis 'H1' has 29 times more support than its complement.
```





# Two methods to calculate the penalty
There are two methods that can be used in calculating the penalty. The default method is often much faster (if the number of parameters is not too high) and needs less input specification. It can, however, not deal with hypotheses that are not of full row-rank (like $H_1$ above). In that case, `restriktor` uses automatically the other (bootstrap) method. In this bootstrap method, one can also more easily change the number of iterations on which the penalty is based (mix.bootstrap). The computation time of this bootstrap method can be reduced by using multiple cores. For a windows device, you then have to use 'parallel = "snow"' (see the tutorial for more options). To use this bootstrap method (on a windows machine), use:
```{r}
if (!require("parallel")) install.packages("parallel") # install this package first (once)
library(parallel)
nrCPUcores <- detectCores(all.tests = FALSE, logical = TRUE)
set.seed(123) # Set seed value
output_b <- goric(lm_fit_Lucas, H0, H1, H2,
                  mix.weights = "boot", parallel = "snow", ncpus = nrCPUcores, mix.bootstrap = 99999)
summary(output_b)
```
This, of course, renders the same results as above (if there is a difference, it is in the second decimal of the penalty).


### Note: not of full row-rank

If the restriction matrix is not of full row-rank, this means one of the following:

a) There is at least one redundant restriction.

   Then, either a.1) leave the redundant one out or a.2) use another (more time-consuming) way of obtaining the level probabilities for the penalty term (goric function does this by default) - as discussed above.

b) There is at least one range restriction (e.g., -2 < group1 < 2). 

   Such a restriction can be evaluated but there is a sensitivity (of a scaling factor in the covariance matrix, like with a prior in a Bayes factor) which currently cannot be checked for.

c) There is at least one conflicting restriction (e.g., 2 < group1 < -2). 

   Such a restriction can evidently never hold and is thus impossible to evaluate. To prevent this type of error delete the one that is incorrect and apply the goric() again.
   


[//]: #The following line is needed to prevent R Markdown from including a lot of white space below the last content.
<div class="tocify-extend-page" data-unique="tocify-extend-page" style="height: 0;"></div>